Diagnosising the sensitivity of grounding line flux to changes in sub-ice shelf melting

### 2019-04-09 draft_v1.tex
go throught the introduction and model description sections

Marine ice sheets like West Antarctica (and to a lesser extent, portions of East Antarctica) are grounded below sea level and their bedrock would remain so even after full isostatic rebound [REFS]. This and the fact that ice sheets generally thicken inland lead to a geometric configuration that is unstable; a small increase in flux at the grounding line thins the ice there, leading to floatation, a retreat of the grounding line into deeper water, further increases in flux (due to thicker ice), and further thinning and grounding line retreat. This theoretical ``marine ice sheet instability'' mechanism [Mercer,Schoof1] is supported by idealized [Schoof2,MISMIP+?] and realistic ice sheet modeling [Gudmundson,others?] experiments and some studies [Rignot, Joughin] argue that such an instability is currently under way along outlet glaciers of Antarctica's Amundsen Sea Embayment (ASE). The relevant perturbation for grounding line retreat in the ASE is thought to be intrusions of relatively warm, intermediate depth ocean waters onto the continental shelves [ref to recent review papers in Oceanography? or other recent reviews?], which have reduced the thickness and extent of marginal ice shelves via increased submarine melting [REFS]. These reductions are critical because fringing ice shelves restrict the flux of ice across their grounding lines farther upstream -- the so-called ``buttressing'' affect of ice shelves [gudmundssonOLD,gudmundsson2013,GudmundssonAndDeRydtPaperOnLarsenC?] -- which makes them a critical control on ice flux from Antarctica to the ocean.

@@

%Warm ocean water is getting more evident of being responsible of increased melt at the base of ice shelves in Antarctica \citep{jenkins2018,rintoul2016}. The basal melting decreases ice thickness and reduce the buttressing ability of ice shelf \citep{gagliardini2010}, and can therefore potentially destabilize marine ice sheets. Satellite altimetry measurements have proved this mechanism that ice shelf thickness decrease links glacier acceleration in Antarctica \citep{pritchard2012}. Recent studies also suggested that the thinning of ice shelfves also contributes to the ice dynamic changes of upstream grounded ice sheets \citep{minchew2018}, indicating the importance of ice shelf-ocean coupling in the whole ice sheet dynamics. 
On ice shelves, gradients in hydrostatic pressure are balanced by the primarily extensional flow of ice towards the calving front \citep{schoof2007}[add a more basic ref, like Paterson, Van der Veen?] and, in theory, a one-dimensional ice shelf provides no buttressing \citep{schoof2007,gudmundsson2013}. \textit{SP: I wonder if this requires a monotonic decrease in ice shelf thickness though?} For realistic, three-dimensional ice shelves however,  buttressing results from three main sources: 1) compressive ice flow 2) lateral shear, and 3) ``hoop'' stress \citep{pegler2012}. \textit{For completeness, we should probably briefly describe how each of these contributes to buttressing rather than assume people already know (?). The first two are easy. I'm not sure about the 3rd.} Due to the complex geometries, kinematics, and dynamics of real ice shelves, an understanding of the specific processes and locations that control ice shelf buttressing is far from straigtforward.

Several recent studies apply whole-Antarctic ice sheet models optimized to present-day observations towards improving our understanding for how Antarctica ice shelves limit flux across the grounding line and, by extension, ice dynamics farther inland. \cite{furst2016} ... \textit{add a few more details on their methods here?} ... to identify regions of the ice shelves that are dynamically ``passive'', such that increased submarine melting, or even complete removal of ice in these areas should not significantly alter local or regional ice dynamics or the flux of ice upstream. \cite{reese2018} used perturbation experiments to link small, localized decreases in ice shelf thickness to changes in integrated grounding line flux (GLF), thereby providing a map of GLF sensitivity to local increases in submarine melt rates. 
%SP: After writing this sentence, it occurs to me that something else that we could follow up on at some point is if / how broader patterns of ice thickness change might complicate the interpretation of these simple perturbation experiments. E.g., you can imagine that 1) multiple perturbations in diff. areas of the ice shelf might conflict with one another and have a more complex impact on g.l. flux (e.g., simultaneous perturbations in multiple areas that show both g.l. flux increases and decreases w/ ice thickness perturbations), 2) similar scale perturbations but applied over wider areas (e.g., over an area that is representative of the actual length scale of submarine melt patterns). Maybe this would be implicit in any 2nd or later study where we perturbed ice shelves based on the sensitivity maps we derive from the adjoint method.

Motivated by these (and other? which?) studies, we build on and extend the methods and analysis of Furst et al. and Reese et al. in order to make progress towards answering the following questions: 
(1) How do local and regional changes in ice shelf geometry affect affect distal changes in GLF? (2) Can local or regional ice shelf dynamics explain GLF sensitivity to local or regional changes in ice shelf thickness? (3) Can we derive and define new tools and analyses for understanding how observed or modeled spatial patterns in submarine melting influence GLF and, by extension, project how changes in submarine melt pattern and magnitude will impact GLF in the future?

@@

Below, we first provide a brief description of the ice sheet model used in our study. We follow with a description of the model experiments and a discussion of the experimental results and their interpretation. We then demonstrate and discuss the pros and cons of a number of possible metrics for quantifying GLF sensitivity to changes in submarine melt. Based on limitations in all metrics explored here, we conclude by proposing and demonstrating an adjoint-based calculation that provides a sensitivity map analogous to that from the \cite{reese2018} perturbation experiments but at the cost of a single model adjoint solve.

@@

Here, we apply MALI to experiments on both idealized and realistic marine-ice sheet geometries. For our idealized domain and model state, we start from the equilibrium initial conditions for the MISMIP+ experiments, as described in \cite{asay2016} and Cornford and others (MISMIP+ papers). The model mesh is spatially uniform at 2 km resolution. For our realistic domain, we use Antarctica's Larsen C ice shelf and its upstream catchement area. The model state is based on the optimization of the ice stiffness ($\gamma$ in Equation \ref{eq:effvisc}) and basal friction ($\beta$ in Equation \ref{eq:basalbc}) coefficients to match present-day velocities \citep{rignot2014} using adjoint-based methods discussed in \cite{hoffman2018} and \cite{perego2014}. The domain geometry is based on BEDMAP2 \citep{fretwell2013} and ice temperatures, which are held fixed for this study, are based on \cite{liefferinge2013}. Mesh resolution coarsens to 20 km in the ice sheet interior and is no greater than 6 km within the ice shelf \textit{This seems coarse to me ... don't we go to finer resolution, e.g. 2 km, near the g.l.?}. Following optimization to present-day velocities, the model is relaxed using a 100 year forward run and it is this initial condition from which the experiments discussed below are conducted. Both the MISMIP+ and Larsen C experiments use 10 vertical layers that are finest near the bed and coarsen towards the surface (4\% and 23\% of the total thickness, respectively). The grounding line position is determined from hydrostatic equilibrium and a sub-element parameterization analogous to method SEP3 from \cite{seroussi2014} is used to define basal friction coefficient values at the grounding line.

>>>>

Marine ice sheets like West Antarctica (and to a lesser extent, portions of East Antarctica) are grounded below sea level and their bedrock would remain so even after full isostatic rebound \citep{barletta2018}. This and the fact that ice sheets generally thicken inland lead to a geometric configuration that is unstable; a small increase in flux at the grounding line thins the ice there, leading to floatation, a retreat of the grounding line into deeper water, further increases in flux (due to thicker ice), and further thinning and grounding line retreat. This theoretical ``marine ice sheet instability'' mechanism \citep{mercer1978, schoof2007} is supported by idealized \citep{schoof2012, asay2016} and realistic ice sheet modeling \citep{royston2016} experiments and some studies \citep{joughin2014,rignot2014} argue that such an instability is currently under way along outlet glaciers of Antarctica's Amundsen Sea Embayment (ASE). The relevant perturbation for grounding line retreat in the ASE is thought to be intrusions of relatively warm, intermediate depth ocean waters onto the continental shelves [ref to recent review papers in Oceanography? or other recent reviews? {\bf{TZ: perhaps Xylar can help with this?}}], which have reduced the thickness and extent of marginal ice shelves via increased submarine melting [REFS]. These reductions are critical because fringing ice shelves restrict the flux of ice across their grounding lines farther upstream -- the so-called ``buttressing'' affect of ice shelves \citep{gudmundsson2012, gudmundsson2013, derydt2015} -- which makes them a critical control on ice flux from Antarctica to the ocean.

@@

%Warm ocean water is getting more evident of being responsible of increased melt at the base of ice shelves in Antarctica \citep{jenkins2018,rintoul2016}. The basal melting decreases ice thickness and reduce the buttressing ability of ice shelf \citep{gagliardini2010}, and can therefore potentially destabilize marine ice sheets. Satellite altimetry measurements have proved this mechanism that ice shelf thickness decrease links glacier acceleration in Antarctica \citep{pritchard2012}. Recent studies also suggested that the thinning of ice shelfves also contributes to the ice dynamic changes of upstream grounded ice sheets \citep{minchew2018}, indicating the importance of ice shelf-ocean coupling in the whole ice sheet dynamics. 
On ice shelves, gradients in hydrostatic pressure are balanced by the primarily extensional flow of ice towards the calving front \citep{hutter1983, morland1987, schoof2007} and, in theory, a one horizontal dimension ($x$-$z$) ice shelf provides no buttressing \citep{schoof2007,gudmundsson2013}. (\textit{SP: I wonder if this requires a monotonic decrease in ice shelf thickness though? \bf{TZ: I think so}}) For realistic, three-dimensional ice shelves however,  buttressing results from three main sources: 1) compressive ice flow 2) lateral shear, and 3) ``hoop'' stress \citep{pegler2012}. Both compressive and lateral shear stresses supply backward resistences to extentional ice shelf flow, and the ``hoop'' stress is a transverse stress arising from azimuthal extension existing in the diverging ice flow regions \citep{wearing2016}. (\textit{For completeness, we should probably briefly describe how each of these contributes to buttressing rather than assume people already know (?). The first two are easy. I'm not sure about the 3rd. \bf{TZ: I add a sentence in the previous line}}) Due to the complex geometries, kinematics, and dynamics of real ice shelves, an understanding of the specific processes and locations that control ice shelf buttressing is far from straigtforward.

Several recent studies apply whole-Antarctic ice sheet models optimized to present-day observations towards improving our understanding for how Antarctica ice shelves limit flux across the grounding line and, by extension, ice dynamics farther inland. \cite{furst2016} calculated the buttressing across Antarctica ice shelves along two major directions (flow and second principle stress) and evaluated their impacts on upstream ice dynamics to identify regions of the ice shelves that are dynamically ``passive'', such that increased submarine melting, or even complete removal of ice in these areas should not significantly alter local or regional ice dynamics or the flux of ice upstream. \cite{reese2018} used perturbation experiments to link small, localized decreases in ice shelf thickness to changes in integrated grounding line flux (GLF), thereby providing a map of GLF sensitivity to local increases in submarine melt rates. 
%SP: After writing this sentence, it occurs to me that something else that we could follow up on at some point is if / how broader patterns of ice thickness change might complicate the interpretation of these simple perturbation experiments. E.g., you can imagine that 1) multiple perturbations in diff. areas of the ice shelf might conflict with one another and have a more complex impact on g.l. flux (e.g., simultaneous perturbations in multiple areas that show both g.l. flux increases and decreases w/ ice thickness perturbations), 2) similar scale perturbations but applied over wider areas (e.g., over an area that is representative of the actual length scale of submarine melt patterns). Maybe this would be implicit in any 2nd or later study where we perturbed ice shelves based on the sensitivity maps we derive from the adjoint method.

%TZ: Yes. This is what I've been thinking of for a while. What I am not quite sure is how we design such complex (but also follow some simplied patterns) perturbation conflicts? Is there anything we can borrow from the ocean modeling results (e.g. ALCC)?

Motivated by these studies, we build on and extend the methods and analysis of \cite{furst2016} and \cite{reese2018} in order to make progress towards answering the following questions: 
(1) How do local and regional changes in ice shelf geometry affect distal changes in GLF? (2) Can local or regional ice shelf dynamics explain GLF sensitivity to local or regional changes in ice shelf thickness? (3) Can we derive and define new tools and analyses for understanding how observed or modeled spatial patterns in submarine melting influence GLF and, by extension, project how changes in submarine melt pattern and magnitude will impact GLF in the future?

@@

Below, we first provide a brief description of the ice sheet model used in our study. We follow with a description of the model experiments and a discussion of the experimental results and their interpretation. We then demonstrate and discuss the pros and cons of a number of possible metrics for quantifying GLF sensitivity to changes in submarine melt. Based on limitations in all metrics explored here, we conclude by proposing and demonstrating an adjoint-based calculation that provides a sensitivity map analogous to that from the \cite{reese2018} perturbation experiments but at the cost of a single model adjoint solve. ({\bf{TZ: we should revisit this paragraph at some point after we finish most of revisions in the manuscript}})

@@

Here, we apply MALI to experiments on both idealized and realistic marine-ice sheet geometries. For our idealized domain and model state, we start from the equilibrium initial conditions for the MISMIP+ experiments, as described in \cite{asay2016} and Cornford and others (MISMIP+ papers) ({\bf{TZ: is it the same paper as Xylar's?}}). The model mesh is spatially uniform at 2 km resolution. For our realistic domain, we use Antarctica's Larsen C ice shelf and its upstream catchement area. The model state is based on the optimization of the ice stiffness ($\gamma$ in Equation \ref{eq:effvisc}) and basal friction ($\beta$ in Equation \ref{eq:basalbc}) coefficients to match present-day velocities \citep{rignot2014} using adjoint-based methods discussed in \cite{perego2014} and \cite{hoffman2018}. The domain geometry is based on BEDMAP2 \citep{fretwell2013} and ice temperatures, which are held fixed for this study, are based on \cite{liefferinge2013}. Mesh resolution coarsens to 20 km in the ice sheet interior and is no greater than 6 km within the ice shelf \textit{This seems coarse to me ... don't we go to finer resolution, e.g. 2 km, near the g.l.?} {\bf{TZ: I think it means the other way around -- the res. is not coarser than 6 km in ice shelf (i.e., $>$2 km and $<$ 6 km)}}. Following optimization to present-day velocities, the model is relaxed using a 100 year forward run and it is this initial condition from which the experiments discussed below are conducted. Both the MISMIP+ and Larsen C experiments use 10 vertical layers that are finest near the bed and coarsen towards the surface (4\% and 23\% of the total thickness, respectively). The grounding line position is determined from hydrostatic equilibrium and a sub-element parameterization analogous to method SEP3 from \cite{seroussi2014} is used to define basal friction coefficient values at the grounding line.
